Perceptions of Inequality and Meritocracy: Their Interplay in Shaping Preferences for Market Justice in Chile (2016-2023)

### 2025-04-29 02-antecedents.qmd
Update 02-antecedents.qmd

---
author: "Equipo EDUMER"
bibliography: "input/bib/merit-edjust.bib"
csl: "input/bib/apa6.csl"
---

# Theoretical views on market justice, inequality perception and meritocracy

@@

Market justice has been discussed in the literature as a normative principle that legitimates the distribution of economic rewards based on individual merit and effort. The most detailed theoretical account of this concept is provided by Lane [-@lane_market_1986], who defines market justice as a system of "earned deserts," whereby individuals are seen as entitled to their outcomes insofar as these are a product of personal productivity and competition. This view directly contrasts with political justice, which emphasizes redistributive mechanisms based on the principles of equality and need, and is often materialized through welfare policies. For Lane [-@lane_market_1986], market justice relies on the conception that markets are neutral and self-regulating arenas, where individuals are treated fairly because they face the same formal rules of engagement. What matters is not whether everyone ends up with the same, but that everyone has the chance to compete, and that success reflects effort and ability. Consequently, the legitimacy of market justice stems from the assumption that inequalities are not only inevitable but fair---so long as the rules are transparent and opportunities are open. These beliefs align with a broader ideological framework that values individual autonomy, private property, and limited government intervention. In this way, market justice provides a moral justification for inequality by framing it as a necessary outcome of personal responsibility and societal efficiency.

Empirical studies have developed different strategies to capture and operationalize market justice preferences. A common approach in the literature is to gauge attitudes toward the legitimacy of inequality in specific domains, especially when linked to income differences. Kluegel and Smith [-@kluegel_beliefs_1981] and later Kluegel et al. [-@kluegel_legitimation_1999] laid the groundwork for measuring beliefs in market justice through survey items that assess whether people support economic inequality when it reflects personal effort and skills. Over time, this approach has been extended beyond income to include other market-mediated outcomes, such as education, healthcare, and/or pensions. For example, @vondemknesebeck_are_2016 and Immergut and Schneider [-@immergut_it_2020] examine whether citizens consider it fair that individuals with higher incomes can access better healthcare, while Lee and Stacey [-@lee_fairness_2023] apply a similar method in the context of education in Australia. These studies usually rely on a survey item asking respondents to evaluate the fairness of income-based access to welfare services, allowing for comparing justice perceptions across different contexts. Some research, like that of Lindh [-@lindh_public_2015] and Svallfors [-@svallfors_political_2007], uses these indicators in combination to measure broader preferences for market-based distribution in welfare systems. More recently, Castillo et al. [-@castillo_socialization_2024] introduced a single-item composite measure of market justice to assess student attitudes toward income-based access to education, healthcare, and pensions in Chile. These different operationalizations all aim to capture the extent to which individuals accept inequality when it is framed as a reflection of market outcomes.

The study of market justice preferences has increasingly focused on how they are shaped by individuals' socioeconomic position, normative beliefs, and the institutional context in which they are embedded. Across the literature, there is strong empirical evidence suggesting that individuals in more advantaged socioeconomic positions---those with higher income, education, and occupational status---are more likely to endorse market-based distribution as fair and legitimate [@svallfors_political_2007; @koos_moral_2019]. This tendency reflects not only material self-interest but also a broader moral economy, in which winners of the market system internalize justifications for the status quo. At the same time, normative and ideological orientations---such as conservative, liberal, or neoliberal values---are consistently associated with greater support for meritocratic principles and skepticism toward redistribution. This is particularly visible in contexts such as Chile and Australia, where studies by Castillo et al. [-@castillo_socialization_2024] and Lee and Stacey [-@lee_fairness_2023] show that those with right-leaning political preferences tend to view income-based access to welfare services as legitimate. Beyond individual-level characteristics, country-level institutions also play a central role. In liberal welfare regimes like those of the United States or the United Kingdom, market justice preferences are more widespread, while in coordinated or social-democratic regimes---such as Sweden or Germany---citizens are generally more critical of market-based inequalities [@immergut_it_2020; @lindh_public_2015]. These findings suggest that market justice is not a fixed belief but a contextually shaped orientation, mediated by structural conditions, cultural values, and the historical development of welfare institutions.

@@

Perceptions of inequality have been associated to attitudes about market justice [@lindh_public_2015; @kluegel_social_1995a]. Research indicates that lower perceived inequality can reinforce support for market-based distributive arrangements by suggesting that the system is fair and that outcomes reflect effort and ability [@kuhn_eye_2011]. In contrast, when inequality is perceived as excessive or structurally determined, individuals could question the legitimacy of market justice and become more supportive of redistributive policies [@castillo_perception_2022; @garcia-sanchez_vicious_2019].

Perceptions of inequality refer to individuals' subjective evaluations of the extent, causes, and consequences of income and wealth disparities. Unlike objective measures such as the Gini index, perceived inequality captures how individuals make sense of distributive hierarchies in their everyday lives, shaped by reference groups, social comparisons, and information environments [@mijs_stratified_2016; @gimpelson_misperceiving_2018; @garcia-castro_perceiving_2020]. Scholars have proposed multiple dimensions of perceived inequality, including its magnitude (how significant are the gaps), vertical structure (between which groups), trend over time (increasing or decreasing), and legitimacy (whether it is just or not) [@engelhardt_what_2018; @garcia-sanchez_vicious_2019]. These dimensions encompass both cognitive and normative aspects of perceptions of inequality and can vary across societies and social groups, depending on exposure, ideology, and personal experience [@castillo_perception_2022; @garcia-sanchez_perceptions_2018].

@@

Meritocracy constitutes a central ideological framework for legitimizing different types od social inequality, for instance through market justice beliefs. Rooted in the belief that rewards and positions should be allocated based on individual effort and talent, meritocracy operates as a normative ideal and a descriptive belief about how society functions. As initially conceptualized by Michael Young [-@young_rise_1962], the term was meant to critique a system in which merit-based stratification becomes a new form of inequality. However, over time, meritocracy has been widely supported in many societies as a fair and desirable principle of distribution, particularly within liberal democracies and market-oriented economies [@mijs_paradox_2021; @sandel_tyranny_2020]. From a sociological perspective, the belief in meritocracy is more than a cognitive assessment; it reflects a moral lens through which individuals interpret inequality. People who believe that success results from hard work and talent are more likely to view social and economic disparities as legitimate [@castillo_percepcion_2019; @batruch_belief_2023]. Conversely, if they see outcomes as driven by luck, social origin, or systemic barriers, inequality is more likely to be perceived as unjust. This distinction becomes crucial in societies with persistent structural inequality, where public narratives often emphasize personal responsibility and merit while overlooking entrenched disadvantages.

@@

These dimensions of meritocracy reflect how respondents perceive society's distributive logic, regardless of whether they personally endorse meritocratic principles. This distinction aligns with recent findings indicating that individuals distinguish between how merit is perceived in society and how it should ideally operate, which in turn shapes their preferences for redistribution and justice [@tejero-peregrina_perceived_2025] Meritocratic beliefs serve as symbolic justifications for unequal outcomes, particularly when access is stratified by income or social opportunity. Prior studies have shown that individuals who perceive higher levels of meritocracy tend to express stronger support for unequal distributions that reflect market outcomes [@castillo_percepcion_2019; @castillo_socialization_2024].

In addition to influencing individual attitudes toward inequality, meritocratic beliefs can contribute to social division and the stigmatization of disadvantaged groups. Recent research has demonstrated that exposure to meritocratic narratives can reinforce the belief that poverty results from individual failings rather than systemic conditions, reducing support for redistributive measures and increasing the stigmatization of the poor [@hoyt_mindsets_2023]. This reinforces negative stereotypes and reduces empathy toward individuals from lower socioeconomic backgrounds. Moreover, @busemeyer_positive_2021 argue that meritocratic narratives can serve as feedback mechanisms that shape public opinion and well-being by framing individuals' understanding of welfare outcomes as deserved or undeserved within existing institutional structures. This psychological mechanism highlights the normative power of meritocracy in stabilizing unequal systems by shaping political attitudes and personal perceptions of success and failure.

@@

First, consistent with previous findings, we expect that a lower perception of economic inequality will be associated with stronger market justice preferences. When individuals perceive smaller economic gaps, they are more likely to view market mechanisms as fair and legitimate, reinforcing acceptance of outcomes based on competition and ability to pay. Conversely, a heightened perception of inequality may erode confidence in market fairness, weakening support for market-based distribution. This relationship is particularly relevant in the context of Chile, where the neoliberal economic model has been a dominant force shaping public attitudes toward inequality and justice.

Second, stronger meritocratic beliefs are expected to be positively associated with market justice preferences. Individuals who believe that effort and talent primarily determine success are more likely to justify unequal outcomes and endorse the notion that markets allocate resources fairly according to individual merit. This aligns with the idea that meritocratic beliefs serve as a moral framework that legitimizes market-based inequalities, as individuals perceive the system as just when they believe that rewards are based on individual merit. This is particularly relevant in contexts where neoliberal ideologies dominate, as they often emphasize individual responsibility and competition as the basis for social order.

Third, we propose that perceptions of meritocracy and perceptions of economic inequality interact in shaping market justice attitudes. Specifically, we argue that the legitimizing effect of meritocratic beliefs on market justice preferences is moderated by perceptions of economic inequality: when perceived inequality is low, the positive association between meritocratic beliefs and market justice preferences will be stronger. However, when perceived inequality is high, this association will weaken, as greater awareness of large economic gaps may challenge the view that outcomes are purely merit-based.

@@

$H1$: Greater perceived economic inequality is associated with stronger market justice preferences.

$H2$: Stronger meritocratic beliefs are associated with stronger market justice preferences.

>>>>

# Theoretical views on market justice, inequality perception, and meritocracy

@@

Conceptually, *market justice* has been discussed in the literature as a normative principle that legitimates the distribution of economic rewards based on individual merit and effort. It is possible to trace the concept to the understanding of Lane [-@lane_market_1986], who defines market justice as a system of "earned deserts," whereby individuals are seen as deserving of a determined distributive outcome due to their effort and skills. In contrast, political justice,  emphasizes principles of equality and need, which are often represented by the welfare state action through social policies. An important remark is that the principles of market justice rely on the assumption that markets are neutral and self-regulating arenas, where individuals are treated fairly because they face the same formal rules of engagement [@lane_market_1986]. Consequently, the legitimacy of market justice stems from the assumption that inequalities are not only inevitable but fair---so long as the rules are transparent and opportunities are open. In this way, market justice provides a moral justification for inequality by framing it as a necessary outcome of individual responsibility.

Empirical studies have shown different strategies for the study of market justice preferences. A common approach in the literature is to gauge attitudes toward the legitimacy of inequality in specific domains, especially when linked to income differences. This can be traced to the seminal work of Kluegel and Smith [-@kluegel_beliefs_1981] who assessed the normative foundations that explain public support for economic inequality. Over time, this approach has been extended beyond income to include other market-mediated outcomes, such as education, healthcare, and/or pensions. For example, @vondemknesebeck_are_2016 and Immergut and Schneider [-@immergut_it_2020] examine whether citizens consider it fair that individuals with higher incomes can access better healthcare, while Lee and Stacey [-@lee_fairness_2023] apply a similar method in the context of education in Australia. These studies usually rely on a survey item asking respondents to evaluate the fairness of income-based access to welfare services, allowing for comparing justice perceptions across different contexts. Similarly, comparative studies have also relied on these indicators to study the support for market-based distribution in welfare systems [@svallfors_political_2007;@lindh_public_2015]. More recently, Castillo et al. [-@castillo_socialization_2024] introduced a single-item composite measure of market justice to assess student attitudes toward income-based access to education, healthcare, and pensions in Chile. These empirical strategies all aim to capture the extent to which individuals accept inequality when framed as a reflection of market outcomes.

The study of market justice preferences has increasingly focused on how they are shaped by individuals' socioeconomic position, normative beliefs, and the institutional context in which they are embedded. Across the literature, there is empirical evidence suggesting that individuals in more advantaged socioeconomic positions---those with higher occupational class, income, and education---are more likely to support market justice principles [@svallfors_political_2007; @koos_moral_2019]. This tendency reflects not only material self-interest but also a broader moral economy, in which winners of the market system internalize justifications for the status quo. At the same time, political ideology also plays a role ---such as economically conservative values ---  where more right-wing individuals show higher support for meritocracy and more skepticism towards redistribution. This is particularly salient in countries with more restricted public provision of social services. For example, Castillo et al. [-@castillo_socialization_2024] in Chile and, Lee and Stacey [-@lee_fairness_2023] in Australia show that those with right-leaning individuals are more supportive of market-based distribution of welfare. Beyond individual characteristics, country-level institutions also play a central role. In liberal welfare regimes like those of the United States or the United Kingdom, market justice preferences are more widespread, while in coordinated or social-democratic regimes---such as Sweden or Germany---citizens are generally more critical of market-based inequalities [@immergut_it_2020; @lindh_public_2015]. These findings suggest that market justice is also shaped contextually, which is represented by the development of welfare institutions.

@@

Perceptions of inequality have been associated with attitudes about market justice [@lindh_public_2015; @kluegel_social_1995a]. Research indicates that lower perceived inequality can reinforce support for market-based distributive arrangements by suggesting that the system is fair and that outcomes reflect effort and ability [@kuhn_eye_2011]. In contrast, when inequality is perceived as excessive or structurally determined, individuals could question the legitimacy of market justice and become more supportive of redistributive policies [@castillo_perception_2022; @garcia-sanchez_vicious_2019].

Perceptions of inequality refer to individuals' subjective evaluations of the extent, causes, and consequences of income and wealth disparities. Unlike objective measures such as the Gini index, perceived inequality captures how individuals make sense of distributive hierarchies in their everyday lives, shaped by reference groups, social comparisons, and information environments [@mijs_stratified_2016; @gimpelson_misperceiving_2018; @garcia-castro_perceiving_2020]. Scholars have proposed multiple dimensions of perceived inequality, including its magnitude (how significant are the gaps), vertical structure (between which groups), the trend over time (increasing or decreasing), and legitimacy (whether it is just or not) [@engelhardt_what_2018; @garcia-sanchez_vicious_2019]. These dimensions encompass both cognitive and normative aspects of perceptions of inequality and can vary across societies and social groups, depending on exposure, ideology, and personal experience [@castillo_perception_2022; @garcia-sanchez_perceptions_2018].

@@

Meritocracy constitutes a central ideological framework for legitimizing different types of social inequality, for instance through market justice beliefs. Rooted in the belief that rewards and positions should be allocated based on individual effort and talent, meritocracy operates as a normative ideal and a descriptive belief about how society functions. As initially conceptualized by Michael Young [-@young_rise_1962], the term was meant to critique a system in which merit-based stratification becomes a new form of inequality. However, over time, meritocracy has been widely supported in many societies as a fair and desirable principle of distribution, particularly within liberal democracies and market-oriented economies [@mijs_paradox_2021; @sandel_tyranny_2020]. From a sociological perspective, the belief in meritocracy is more than a cognitive assessment; it reflects a moral lens through which individuals interpret inequality. People who believe that success results from hard work and talent are more likely to view social and economic disparities as legitimate [@castillo_percepcion_2019; @batruch_belief_2023]. Conversely, if they see outcomes as driven by luck, social origin, or systemic barriers, inequality is more likely to be perceived as unjust. This distinction becomes crucial in societies with persistent structural inequality, where public narratives often emphasize personal responsibility and merit while overlooking entrenched disadvantages.

@@

These dimensions of meritocracy reflect how respondents perceive society's distributive logic, regardless of whether they endorse meritocratic principles. This distinction aligns with recent findings indicating that individuals distinguish between how merit is perceived in society and how it should ideally operate, which in turn shapes their preferences for redistribution and justice [@tejero-peregrina_perceived_2025] Meritocratic beliefs serve as symbolic justifications for unequal outcomes, particularly when access is stratified by income or social opportunity. Prior studies have shown that individuals who perceive higher levels of meritocracy tend to express stronger support for unequal distributions that reflect market outcomes [@castillo_percepcion_2019; @castillo_socialization_2024].

In addition to influencing individual attitudes toward inequality, meritocratic beliefs can contribute to social division and the stigmatization of disadvantaged groups. Recent research has demonstrated that exposure to meritocratic narratives can reinforce the belief that poverty results from individual failings rather than systemic conditions, reducing support for redistributive measures and increasing the stigmatization of the poor [@hoyt_mindsets_2023]. This reinforces negative stereotypes and reduces empathy toward individuals from lower socioeconomic backgrounds. Moreover, @busemeyer_positive_2021 argues that meritocratic narratives can serve as feedback mechanisms that shape public opinion and well-being by framing individuals' understanding of welfare outcomes as deserved or undeserved within existing institutional structures. This psychological mechanism highlights the normative power of meritocracy in stabilizing unequal systems by shaping political attitudes and personal perceptions of success and failure.

@@

First, consistent with previous findings, we expect that a lower perception of economic inequality will be associated with stronger market justice preferences. When individuals perceive smaller income gaps, they are more likely to view market mechanisms as fair and legitimate, reinforcing the acceptance of outcomes based on competition and ability to pay. Conversely, a heightened perception of inequality may erode confidence in market fairness, weakening support for market-based distribution. This relationship is particularly relevant in the context of Chile, where the neoliberal economic model has been a dominant force in shaping public attitudes toward inequality and justice.

Second, higher perceived meritocracy is expected to be positively associated with market justice preferences. Individuals who believe that effort and talent primarily determine success are more likely to justify unequal outcomes and endorse the notion that markets allocate resources fairly according to individual merit. This aligns with the idea that meritocratic beliefs serve as a moral framework that legitimizes market-based inequalities, as individuals perceive the system as just when they believe that rewards are based on individual merit. This is particularly relevant in contexts where neoliberal ideologies dominate, as they often emphasize individual responsibility and competition as the basis for social order.

Third, we propose that perceptions of meritocracy and perceptions of economic inequality interact in shaping market justice attitudes. Specifically, we argue that the legitimizing effect of perceived meritocracy on market justice preferences is moderated by perceived economic inequality: when perceived inequality is low, the positive association between meritocratic beliefs and market justice preferences will be stronger. However, when perceived inequality is high, this association will weaken, as greater awareness of large economic gaps may challenge the view that outcomes are purely merit-based.

@@

$H1$: Higher perceived economic inequality is associated with higher market justice preferences.

$H2$: Higher meritocratic beliefs are associated with higher market justice preferences.
